---
title: "Example 6: Two arm | Fixed vs group-sequential design | Continuous | design comparison"
categories: [two-arm, group-sequential, fixed design, continuous, design comparison, operating characteristics]
description: "A design table mapped with purrr::pmap compares a fixed single-look design against a 2-look O'Brien-Fleming group-sequential design on type-I error, power, and expected sample size."
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  cache    = TRUE
)
```

```{r setup}
library(rxsim)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
```

A fixed single-look design and a group-sequential (GS) design with an early-stopping
rule achieve similar type-I error and power, but the GS design can stop early under
the alternative, spending fewer patients on average. This example compares the two
side-by-side using a **design table** (a `tibble` where each row is a design) mapped
with `purrr::pmap`. The approach enumerate multiple trial strategies, and scales 
naturally to more designs or scenarios.

**Unique focus:** design-table pattern, `purrr::pmap` over a scenario grid, 2-look
O'Brien-Fleming group-sequential, operating characteristics (type-I, power, expected N).

## Design table

Each row in `designs` specifies one trial design: the look fractions and the
corresponding nominal two-sided p-value boundary at each look.

```{r design-table}
# OBF K=2, equal info (0.5, 1.0), two-sided alpha = 0.05.
# Swap gsDesign::gsDesign() for other spending functions.
designs <- tibble(
  design = c("Fixed", "GS-OBF"),
  looks  = list(1.0, c(0.5, 1.0)),
  bounds = list(0.05, c(0.00517, 0.04800))
)
designs
```

`Fixed` performs a single final analysis at full enrollment (`fraction = 1.0`) with
two-sided `alpha = 0.05`. `GS-OBF` adds an interim at 50% information with
O'Brien-Fleming spending: the interim bound is tight (`p < 0.00517`), spending
almost no alpha early, while the final bound (`p < 0.04800`) is only fractionally
narrower than the fixed design's. Together the two bounds preserve overall `α ≈ 0.05`.

## Scenario

```{r scenario}
sample_size <- 100
arms        <- c("pbo", "trt")
allocation  <- c(1, 1)
n_reps      <- 1000   # replicates per scenario row

# Block accrual: 10 subjects per time unit.
# Interim (50 enrolled) fires at t = 5; final (100 enrolled) at t = 10.
enroll_fn  <- function(n) { a <- ceiling(seq_len(n) / 10); c(a[1], diff(a)) }
dropout_fn <- function(n) rep(0, n)

# Null (no effect) and alternative (delta = 0.5 SD): design comparison, not a sweep.
scenarios <- expand_grid(design = designs$design, delta = c(0, 0.5)) |>
  left_join(designs, by = "design") |>
  mutate(seed = 1000L + row_number())
scenarios
```

`expand_grid` produces one row per (design × truth) combination; `left_join` attaches
the `looks` and `bounds` list-columns from the design table. A `seed` column makes
each run reproducible independently.

## Populations

```{r populations}
mk_pops <- function(delta) list(
  pbo = function(n) data.frame(id = 1:n, value = rnorm(n),        readout_time = 0),
  trt = function(n) data.frame(id = 1:n, value = rnorm(n, delta), readout_time = 0)
)
```

## Analysis generators

`mk_gens` builds one named analysis per look fraction. Each fires a t-test on all
subjects enrolled so far.

```{r analysis-generators}
mk_gens <- function(looks) {
  gens <- lapply(looks, function(fr) list(
    trigger  = enroll_trigger(fr, sample_size),
    analysis = function(df, current_time) {
      e <- subset(df, !is.na(enroll_time))
      data.frame(frac = fr, n = nrow(e), p = t.test(value ~ arm, data = e)$p.value)
    }
  ))
  names(gens) <- paste0("look_", seq_along(looks))
  gens
}
```

`enroll_trigger(fr, sample_size)` fires once when enrolled count reaches
`fr × sample_size`. The GS design registers two conditions (`look_1` at n = 50,
`look_2` at n = 100); the Fixed design registers one (`look_1` at n = 100).

## Simulate

`run_one` runs `n_reps` replicates for one scenario row. After collecting results it
applies the per-look boundary vector element-wise, `p[k] < bounds[k]`, to find
the first look where the trial crosses its threshold. `n_used` is the sample size
at that look, or the full 100 if no boundary is crossed.

```{r simulate, cache=TRUE}
run_one <- function(design, delta, looks, bounds, seed) {
  set.seed(seed)
  tr <- replicate_trial(
    design, sample_size, arms, allocation, enroll_fn, dropout_fn,
    mk_gens(looks), mk_pops(delta), n = n_reps
  )
  invisible(run_trials(tr))

  per <- collect_results(tr) |>
    arrange(replicate, frac) |>
    group_by(replicate) |>
    summarise(
      first_rej = { i <- which(p < bounds); if (length(i)) min(i) else NA_integer_ },
      n_used    = if (is.na(first_rej)) last(n) else n[first_rej],
      .groups   = "drop"
    )

  tibble(
    design     = design,
    scenario   = if (delta == 0) "null (d=0)" else sprintf("alt (d=%.1f)", delta),
    reject_h0  = mean(!is.na(per$first_rej)),
    expected_n = mean(per$n_used)
  )
}

oc <- pmap(scenarios, run_one) |> list_rbind()
```

For the Fixed design `bounds` is the scalar `0.05`, so `which(p < 0.05)` on a
single p-value. For GS-OBF `bounds` is `c(0.00517, 0.04800)` and `p` is a length-2
vector, so the comparison is element-wise: interim bound at look 1, final bound at
look 2.

## Results

```{r results}
oc
```

Both designs control type-I error near 0.05 and reach similar power under
`delta = 0.5`. The GS design's expected N drops below 100 under the alternative
because some replicates stop at the interim when early evidence is compelling.

```{r self-check, include=FALSE}
stopifnot(
  oc$reject_h0[oc$design == "Fixed"   & oc$scenario == "null (d=0)"] < 0.15,
  oc$expected_n[oc$design == "GS-OBF" & grepl("alt", oc$scenario)]   < 100
)
```

## Operating characteristics

```{r plot-power}
ggplot(oc, aes(design, reject_h0, fill = scenario)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0.05, linetype = 3, colour = "grey50") +
  coord_cartesian(ylim = c(0, 1)) +
  scale_fill_manual(values = c("#e18600", "#076d7e")) +
  labs(
    x     = NULL,
    y     = "P(reject H0)",
    fill  = NULL,
    title = "Type-I error and power: fixed vs group-sequential"
  )
```

```{r table-oc}
knitr::kable(
  oc,
  digits  = c(0, 0, 3, 1),
  col.names = c("Design", "Scenario", "P(reject H0)", "Expected N"),
  caption = "Operating characteristics: type-I error, power, and expected sample size."
)
```

The power chart confirms both designs control type-I error near 0.05 and deliver
comparable power. The table shows the GS design's payoff: expected N drops below
100 under the alternative because some replicates stop at the interim when early
evidence is compelling, while the Fixed design always uses all 100 patients.

## Next steps

- [Example 1](example-1.html): the minimal fixed design this GS design extends
- [Example 4](example-4.html): Bayesian Go/No-Go, a different decision framework
- [Conditions and Triggers](https://boehringer-ingelheim.github.io/rxsim/articles/conditions.html): multi-look designs in depth
