---
title: "Example 2: Two arm | Fixed design | Two correlated continuous endpoints | t-test"
categories: [two-arm, fixed design, continuous, correlated endpoints, multiplicity, t-test]
description: "Correlated endpoints via `MASS::mvrnorm`, Holm correction, joint vs marginal rejection rates as correlation varies."
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
# Core package that provides Timer, Population, Trial, deterministic_schedule, add_timepoints, ...
library(rxsim)

# Utilities
library(dplyr)
library(MASS)   # for mvrnorm (simulate correlated endpoints)
library(ggplot2)
```

Many trials evaluate co-primary or key secondary endpoints simultaneously.
Ignoring endpoint correlation when planning a study leads to incorrect power
estimates. This example shows how to simulate two correlated continuous
endpoints (e.g., a primary efficacy score and a key secondary biomarker) and
apply multiplicity adjustment via Holm's procedure. See [Example
1](example-1.html) for the simpler single-endpoint baseline.

**Unique focus:** correlated endpoints via `MASS::mvrnorm`, Holm correction,
joint vs marginal rejection rates as correlation varies.

Common scaffolding (packages, scenario declaration, and `replicate_trial()`
pattern) follows [Example 1](example-1.html); this vignette only expands the
parts that change for correlated multi-endpoint analyses.

## Scenario

We consider a **two-arm, fixed design** with **two correlated continuous
endpoints** per subject. A single analysis fires at full enrollment.

```{r scenario}
# Total target sample size
sample_size <- 100

# Arms and allocation (balanced)
arms       <- c("pbo", "trt")
allocation <- c(1, 1)

# Endpoint model parameters
# Endpoint correlation structure (common across arms for simplicity)
rho  <- 0.60            # correlation between endpoint1 and endpoint2
sd1  <- 1.00            # SD of endpoint1
sd2  <- 1.00            # SD of endpoint2

# Mean structure per arm
mu1_pbo <- 0.00         # Control mean for endpoint1
mu2_pbo <- 0.00         # Control mean for endpoint2

# Treatment effects (mean shifts)
delta1  <- 0.65         # Treatment - control difference for endpoint1
delta2  <- 0.60         # Treatment - control difference for endpoint2

mu1_trt <- mu1_pbo + delta1
mu2_trt <- mu2_pbo + delta2

# Construct covariance matrix
Sigma <- matrix(
  c(sd1^2, rho*sd1*sd2,
    rho*sd1*sd2, sd2^2),
  nrow = 2, byrow = TRUE
)

scenario <- tidyr::expand_grid(
  sample_size = sample_size,
  allocation = list(allocation),
  rho = rho,
  delta1 = delta1,
  delta2 = delta2
)

enrollment_fn <- function(n) c(1, rep(0, n - 1))  # all enroll at t = 1 (single look)
dropout_fn    <- function(n) rep(0, n)            # no dropout
```

`rho = 0.60` represents a moderate positive correlation between the two
endpoints - plausible when both reflect the same underlying biological process.
The covariance matrix `Sigma` is constructed from the marginal SDs and
correlation using the standard identity cov(y1, y2) = rho × sd1 × sd2. Treatment
shifts are delta1 = 0.65 SD for the primary endpoint and delta2 = 0.60 SD for
the secondary, reflecting a scenario where the primary endpoint is more
sensitive to treatment.


## Populations

For each arm we simulate **two correlated endpoints** per subject using
`MASS::mvrnorm`. In this vignette, that arm-specific generator is the
**custom data generating model** for endpoint behavior.

```{r population}
#| code-fold: true
#| code-summary: "Correlated endpoint population generators for placebo and treatment arms"
# Create closure to capture parameters
mk_population_generator <- function(mu_y1, mu_y2) {
  function(n) {
    xy <- MASS::mvrnorm(
      n     = n,
      mu    = c(mu_y1, mu_y2),
      Sigma = Sigma
    )
    data.frame(
      id = seq_len(n),
      y1 = xy[, 1],
      y2 = xy[, 2],
      readout_time = 1
    )
  }
}

population_generators <- list(
  pbo = mk_population_generator(mu1_pbo, mu2_pbo),
  trt = mk_population_generator(mu1_trt, mu2_trt)
)
```

`MASS::mvrnorm` draws `n` samples from a multivariate normal with the given mean
vector and covariance matrix `Sigma`. The closure pattern
(`mk_population_generator`) captures the arm-specific means at definition time,
so each call to `population_generators$pbo(n)` or `population_generators$trt(n)`
produces data under the correct arm parameters. Endpoint correlation flows
entirely through the shared `Sigma` matrix - changing `rho` in the scenario
automatically updates both arms.

## Conditions

This is a **fixed design**: perform analysis **once**.
We run two independent two-sample t-tests (one per endpoint).

```{r trigger}
analysis_generators <- list(
  final = list(
    trigger = enroll_trigger(1.0, sample_size),
    analysis = function(df, current_time) {
      df_e <- df |> subset(!is.na(enroll_time))

      p1 <- t.test(y1 ~ arm, data = df_e)$p.value
      p2 <- t.test(y2 ~ arm, data = df_e)$p.value
      padj <- p.adjust(c(p1, p2), method = "holm")

      data.frame(
        scenario,
        p_y1      = unname(p1),
        p_y2      = unname(p2),
        p_holm_y1 = unname(padj[1]),
        p_holm_y2 = unname(padj[2]),
        n_total   = nrow(df_e),
        n_pbo     = sum(df_e$arm == "pbo"),
        n_trt     = sum(df_e$arm == "trt"),
        stringsAsFactors = FALSE
      )
    }
  )
)
```

The two t-tests are run independently on the enrolled subset, producing
unadjusted p-values `p_y1` and `p_y2`. `p.adjust(..., method = "holm")` applies
Holm's stepwise procedure. `padj[1]` and `padj[2]` stay in endpoint order
(`y1`, `y2`) and account for familywise error across both endpoints, so each
adjusted value is never smaller than its unadjusted counterpart.

## Simulate

```{r trial, results=FALSE}
set.seed(3)
trials <- replicate_trial(
  trial_name            = "two_endpoints_fixed_design",
  sample_size           = sample_size,
  arms                  = arms,
  allocation            = allocation,
  enrollment            = enrollment_fn,
  dropout               = dropout_fn,
  analysis_generators   = analysis_generators,
  population_generators = population_generators,
  n                     = 5
)
run_trials(trials)
```

## Results

One row per replicate, row-bound across all replicates:

```{r results}
replicate_results <- collect_results(trials)
replicate_results
```

`collect_results()` prepends `replicate`, `timepoint`, and `analysis` columns to
the four p-value columns. `p_holm_y1` and `p_holm_y2` will always be >= their
unadjusted counterparts because Holm correction is more conservative. Because
the two endpoints are positively correlated (rho = 0.6), replicates that reject
for y1 tend to also reject for y2 - joint rejection probability exceeds what you
would expect for independent endpoints under the same effect sizes.

## Power curve

Higher endpoint correlation affects the joint rejection rate in a
counter-intuitive way: positive correlation means both endpoints tend to move
together, so a trial that is "lucky" in one endpoint is often lucky in the other.
The following sweep quantifies marginal and joint power across four `rho` values.

```{r power_corr, cache=TRUE}
#| code-fold: true
#| code-summary: "Power sweep across endpoint correlations (rho = 0.0, 0.3, 0.6, 0.9)"
set.seed(43)
n_reps_pw <- 200
rho_vals <- c(0.0, 0.3, 0.6, 0.9)

pw_df3 <- do.call(rbind, lapply(rho_vals, function(r) {
  Sig_r <- matrix(c(1, r, r, 1), 2)
  pop_pw <- list(
    pbo = local({
      S <- Sig_r
      function(n) {
        xy <- MASS::mvrnorm(n, c(0, 0), S)
        data.frame(id = 1:n, y1 = xy[, 1], y2 = xy[, 2], readout_time = 1)
      }
    }),
    trt = local({
      S <- Sig_r
      function(n) {
        xy <- MASS::mvrnorm(n, c(delta1, delta2), S)
        data.frame(id = 1:n, y1 = xy[, 1], y2 = xy[, 2], readout_time = 1)
      }
    })
  )
  an_pw <- list(final = list(
    trigger  = enroll_trigger(1.0, sample_size),
    analysis = function(df, ct) {
      d <- subset(df, !is.na(enroll_time))
      p1 <- t.test(y1 ~ arm, data = d)$p.value
      p2 <- t.test(y2 ~ arm, data = d)$p.value
      pa <- p.adjust(c(p1, p2), method = "holm")
      data.frame(p1 = p1, p2 = p2, pa1 = pa[1], pa2 = pa[2])
    }
  ))
  tr <- replicate_trial(
    "pw3", sample_size, arms, allocation,
    enrollment_fn, dropout_fn, an_pw, pop_pw, n_reps_pw
  )
  invisible(run_trials(tr))
  res <- collect_results(tr)
  data.frame(
    rho           = r,
    power_y1      = mean(res$p1  < 0.05),
    power_y2      = mean(res$p2  < 0.05),
    power_joint   = mean(res$pa1 < 0.05 & res$pa2 < 0.05)
  )
}))
```

```{r plot_power3}
pw_long3 <- tidyr::pivot_longer(
  pw_df3,
  c(power_y1, power_y2, power_joint),
  names_to = "metric", values_to = "power"
)
pw_long3$metric <- factor(
  pw_long3$metric,
  levels = c("power_y1", "power_y2", "power_joint"),
  labels = c("y1 (marginal)", "y2 (marginal)", "Joint (Holm)")
)

ggplot(pw_long3, aes(rho, power, colour = metric)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.80, linetype = 2, colour = "grey50") +
  scale_colour_manual(values = c("#e18600", "#076d7e", "#00E47C")) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Endpoint correlation rho",
    y = "Empirical power (alpha = 0.05)",
    title = "Marginal and joint power vs endpoint correlation",
    colour = NULL
  )
```

## Next steps

- [Example 3](example-3.html) - adds a time-to-event endpoint alongside
  the continuous one
- [Population](https://boehringer-ingelheim.github.io/rxsim/articles/class-population.html) - endpoint data setup for continuous,
  binary, and time-to-event outcomes
- [Enrollment and Dropout](https://boehringer-ingelheim.github.io/rxsim/articles/enrollment.html) - piecewise enrollment
  with `deterministic_schedule()`
